---
title: "Military Spending vs GDP Dashboard"
author: "Trevor Sully, Balaji Prasanna Venkatesh, Mitch Kim"
format:
  dashboard:
    theme: lux
    logo: "iSchoolSymbolOnlyRGB_2024LightGray.svg"
    fontsize: 1.002em
    embed-resources: true
---

# Project Overview

**Project Introduction**\

Military spending reflects the amount of currency a country will allocate towards its budget for defense and security. In contrast, the GDP per capita is a measure of the economic freedom and spending power the average resident owns. With recent conflicts, many countries have raised or plan to up their defense budgets due to geopolitical factors. In a recent event, Israel and the United States of America have attacked Iran, leading to an increase in American military spending. In an article from [Fortune](https://fortune.com/2026/03/02/how-much-trump-iran-war-operation-epic-fury-cost-taxpayers/), it is written “According to Kent Smetters, director of the Penn Wharton Budget Model (PWBM) and one of the nation’s foremost fiscal analysts, the total economic cost of the strikes could reach as high as \$210 billion.” This raises questions on how increases in the military budget affects countries economically through its residents. 

### Big picture questions

**Big picture question 1**\
How do the top spenders on military budgets compare to the lowest in their GDP per capita, and do the wealthier countries spend more on their budget? 

**Big picture question 2**\
Does a larger defense and security budget create more economic prosperity through jobs, or does the defense budget struggle to reach citizens?

**Big picture question 3**\
Do countries that change their military budget over time, either by raising or lowering it, see differences in their GDP per capita?

### How should we answer these questions

To answer these questions, we found and compared two data sets of GDP per capita and military spending. The Stockholm International Peace Institute ([SIPRI](https://www.sipri.org/databases/milex)) is the creator of our data set for military spending overtime. This index has the military spending of countries since 1949, with the total amount spent on military in a year representing the amount for each calendar year. Our data set for GDP per capita comes from the World Bank Group ([World Bank](https://www.worldbank.org/ext/en/home)).

This is an important question to ask with border disputes and different geopolitical factors in today’s society. In recent conflicts, the United States of America has recently used its vast military budget in conflicts in Venezuela and Iran. As this happens, many Americans worry about if this is a good use of our money as many economies start to dwindle in recent years. With other wars and invasions across the world, the economic impact is important to learn about with examples like Russia and Ukraine, specifically how the citizens of these countries are affected other than war.

```{r}
#| warning: false
library(tidyverse)
library(ggplot2)
library(plotly)
```

```{r}
# Read the first CSV
military <- read.csv("military-spending_share_of_GDP.csv")

# Read the second CSV
gdp <- read.csv("gdp-per-capita.csv")

# Joining two CSV file together
data <- left_join(military, gdp, by = c("Entity", "Year"))

write.csv(data, "military_gdp.csv", row.names = FALSE)

```

# Top 20 GDP per Capita

**Graph 1: Top 20 GDP per Capita(1990 - 2024)**\

**Q: How do countries among the top of GDP per capita compare to each other?**\

A majority of these countries experience around the same GDP per capita at around 50,000 USD. The 8 countries among the top of GDP per capita are outside of this range, with Luxembourg and Qatar exceeding 100,000 USD.

```{r}
top20 <- data |>
  group_by(Entity) |>
  summarize(
    avg_gdp = mean(GDP.per.capita, na.rm = TRUE),
    avg_military = mean(Military.expenditure....of.GDP., na.rm = TRUE)
  ) |>
  arrange(desc(avg_gdp)) |>
  slice(1:20)

graph1 <- ggplot(top20, aes(x = reorder(Entity, avg_gdp), y = avg_gdp)) +
    geom_col() +
    coord_flip() +
    labs(
      title = "Top 20 Countries by Average GDP per Capita",
      x = "Country",
      y = "Average GDP per Capita ($)"
    ) 

ggplotly(graph1)
```

# Top 20 Military Spending (% of GDP)

**Graph 2: Average Military Spending (% GDP)(1990 - 2024)**\

**Q: How much of the budget among these countries is portioned to military expenditure?**\

This graph has more variation in the results, with some countries spending as little as 1 percent of the national GDP, while the top countries spend close to 10 percent. This graph also seems to correlate region with military spending. Many European countries prioritize relatively little of their national budget towards military expenditure, while middle eastern nations seem to have an overall higher budget for security.

```{r}
graph2 <- ggplot(top20, aes(x = reorder(Entity, avg_military), y = avg_military)) +
    geom_col() +
    coord_flip() +
    labs(
      title = "Average Military Spending (% of GDP) for Top 20 GDP Countries",
      x = "Country",
      y = "Average Military Spending (% GDP)"
    )

ggplotly(graph2)
```

# GDP per Capita Over Time for Top 20 Countries

**Q: Graph 3: GDP per Capita Over Time for Top 20 Countries**\

**How has the GDP per capita evolved over time for these nations?**\

Many of these countries have experienced consistent economic growth over time with a few exceptions like Brunei and the United Arab Emirates. This shows that most countries among the top military spenders have grown their GDP per capita in recent decades.

```{r}
top20 <- data |>
  group_by(Entity) |>
  summarize(
    avg_gdp = mean(GDP.per.capita, na.rm = TRUE),
    avg_military = mean(Military.expenditure....of.GDP., na.rm = TRUE)
  ) |>
  arrange(desc(avg_gdp)) |>
  slice(1:20)

top20_data <- data |>
  filter(Entity %in% top20$Entity) |>
  filter(Year >= 1990)

graph3 <- ggplot(top20_data, aes(
    x = Year,
    y = GDP.per.capita,
    color = Entity)) +
  geom_line() +
  labs(
    title = "GDP per Capita Over Time for Top 20 Countries",
    x = "Year",
    y = "GDP per Capita ($)",
    color = "Country"
  )

ggplotly(graph3)
```

# Conclusion

**Conclusion**\

**The first major takeaway from the graphs we created for these two datasets is that the countries with the highest GDP per capita do not always spend a large share on their own military.** Countries such as Luxembourg, Norway, and Ireland utilize a relatively small military budget and contain a high GDP per capita with mostly consistent growth GDP wise since the 1990’s. In contrast, there are many countries that spend a larger percent on military but still place near the top on GDP per capita. Countries like Saudi Arabia and Kuwait are at the top of military spending and still have a large GDP per capita. It should be noted that these countries have a lower GDP per capita compared to others that spend less on their military.

**A second takeaway is that geopolitical factors seem to play a large part in a country's military expenditures.** Many of the nations at the top of military spending are middle eastern countries like Saudi Arabia, Kuwait, Qatar, and the United Arab Emirates. This suggests that regional conflicts and security play a larger influence on military spending. This analysis concludes that wealthier countries do not need to spend more on military expenditures and that the relationship with a nation's security budget is complex.

**The third takeaway from our research is that many of these countries have a consistent GDP growth rate over time.** A couple of outliers to this are Brunei and Qatar, which in the last decade have been over a consistent decline for GDP. This suggests that a military spending budget can exceed a beneficial amount for a country and start to decline the financial benefit for its citizens. It is noted that these countries may have certain political factors that play a part into their own military expenditure. The growth of GDP also varies from military spending, concluding that as a country experiences financial growth they may prioritize other areas of spending, rather than just military.

**Overall**, our data comes from unbiased sources that accurately represent the relationships between military spending and GDP per capita. There should not be any inaccurate estimations, but our findings may exclude countries that are relatively new or do not prioritize military spending. In the future, including more baseline countries into the data may lead to more conclusions for our relationship between the two factors. Using a map of regions in the future may also be helpful to describe this relationship, as participation in different regional conflicts seems to play a big factor with the dataset we used.

**In conclusion**, we found that the top spenders of the military are not always among the highest of GDP per capita, whereas many countries with lower budgets maintain a more consistent growing GDP over time. The graph also suggests that military spending seems to correlate with economic prosperity, but a country that overspends will overall hurt their economic trend over time. The graph also answers the question of whether changing military budgets affects the GDP per capita growth over time, showing that this seems to have little correlation on the overall economic growth over time.

# Data Context

**Data Context:**\

-   Military Spending Data Set–https://www.sipri.org/databases/milex Created by SIPRI For this data set, the information is collected by the Stockholm International Peace Institute, where it is military expenditure by using government budgets, defense reports, and international financial statistics. The index has hundreds of records as each entry is for a specific year and contains most countries in the world. For our graphs, the data is from recent years, (1990 to 2025) and covers nations from across the entire world. Because the dataset is broad in its coverage, it can show how different regions spend military budgets differently. A problem with the dataset however is that the numbers are self reported, meaning that some countries may not have reported accurately.

-   GDP per Capita Dataset-https://www.worldbank.org/ext/en/home Created by WorldBank For this dataset, the GDP per capita was collected of most countries by using the economic output of a nation divided by the population. This data was collected over the course of many years, so there are hundreds of entries for this dataset as well. For our graphs, this dataset also is in the time range of 1990 to 2025. This index is also wide and broad, allowing us to examine many different regions and economic situations. Where this dataset falls short is that it does not account for wealth inequalities within countries, meaning that a high GDP per capita on this graph does not always mean a well balanced economy.
